$R^*$: A robust MCMC convergence diagnostic with uncertainty using gradient-boosted machines

### 2020-03-27 src/s_trends.Rmd
added in trends examples

```{r}
res <- readRDS("../output/experiment_trends_all_dim.rds")
ggplot(data = res, aes(y = Rstar_split, x = trend)) + 
  geom_point() + 
  geom_jitter() + 
  facet_grid(. ~ iters) + 
  geom_hline(yintercept = 1) + 
  ggtitle('Split-R*')
```

@@

```{r}
res <- readRDS("../output/experiment_trends_one_dim.rds") %>% 
  select(trend, iters, rep, dimension, Rstar) %>% 
  unique()
ggplot(data = res, aes(x = trend, y = Rstar)) + 
  geom_point() + 
  geom_jitter() + 
  facet_grid(vars(iters), vars(dimension)) + 
  geom_hline(yintercept = 1) + 
  ggtitle('R* without splitting')
```

```{r}
res <- readRDS("../output/experiment_trends_one_dim.rds") %>% 
  select(trend, iters, rep, dimension, Rstar_split) %>% 
  unique()
ggplot(data = res, aes(x = trend, y = Rstar_split)) + 
  geom_point() + 
  geom_jitter() + 
  facet_grid(vars(iters), vars(dimension)) + 
  geom_hline(yintercept = 1) + 
  ggtitle('Split-R*')
```

@@

ggplot(data = res, aes(x = rho, y = Rstar)) + 
  geom_jitter() + 
  facet_grid(vars(iters)) + 
  geom_hline(yintercept = 1) + 
  ggtitle('R* without splitting') +
  ylim(0, 6)

>>>>

```{r}
res <- readRDS("../output/experiment_trends_all_dim.rds")
g <- 
  res %>%
  select(iters, trend, Rstar, Rstar_split) %>% 
  melt(id.vars=c("iters", "trend")) %>% 
  mutate(variable=case_when(variable=="Rstar"~"A. R*",
                            variable=="Rstar_split"~"B. split-R*")) %>% 
  ggplot(aes(y = value, x = trend)) + 
  geom_point() + 
  geom_jitter() + 
  facet_grid(vars(iters), vars(variable)) + 
  geom_hline(yintercept = 1) +
  theme(strip.text.x = element_text(size=12, colour="black"),
        strip.text.y = element_text(size=12, colour="black"),
        axis.text.x = element_text(size=10, colour="black"),
        axis.text.y = element_text(size=10, colour="black")) +
  xlab("Trend") +
  ylab("Value")
ggsave("../output/trends_all_dim.pdf", g, width = 8, height = 6)
```

@@

```{r}
g <- readRDS("../output/experiment_trends_one_dim.rds") %>% 
  select(trend, iters, rep, dimension, Rstar_split) %>% 
  unique() %>%
  select(-rep) %>% 
  melt(c("trend", "iters", "dimension")) %>% 
  filter(variable=="Rstar_split") %>% 
  ggplot(aes(y = value, x = trend)) + 
  geom_point() + 
  geom_jitter() + 
  facet_grid(vars(iters), vars(dimension)) + 
  geom_hline(yintercept = 1) +
  theme(strip.text.x = element_text(size=12, colour="black"),
        strip.text.y = element_text(size=12, colour="black"),
        axis.text.x = element_text(size=10, colour="black"),
        axis.text.y = element_text(size=10, colour="black")) +
  xlab("Trend") +
  ylab("Value")
g

ggsave("../output/trends_one_dim.pdf", g, width = 8, height = 6)
```

# Trend in the joint distribution over time
```{r}
library(mvtnorm)
rmvrnorm2D <- function(n, mux, muy, sigmax, sigmay, rho){
  return(rmvnorm(n, c(mux, muy),
                 matrix(c(sigmax^2, sigmax * sigmay * rho,
                          sigmax * sigmay * rho, sigmay^2),
                        ncol = 2)))
}

conds <- expand.grid(
  iters = c(250, 1000, 4000), 
  rep = 1:10,
  rho = c(0, 0.2, 0.4, 0.6, 0.8)
)
res <- vector("list", nrow(conds))
chains = 4

for (i in 1:nrow(conds)) {
  iters <- conds[i, "iters"]
  rep <- conds[i, "rep"]
  rho <- conds[i, "rho"]
  rho_trend <- seq(-rho, rho, length.out = iters)
  r <- array(dim=c(iters, chains, 2))
  for(j in 1:chains)
    for(jj in 1:iters)
      r[jj, j, ] <- rmvrnorm2D(1, 0, 0, 1, 1, rho_trend[jj])
  # take max as just using the Rhat values
  rs <- as.data.frame(monitor_extra(r)) %>% 
    apply(2, max)

 r_2d <- matrix(nrow = iters * chains, ncol=(2 + 2))
  k <- 1
  for(ii in 1:iters){
    for(j in 1:chains){
      r_2d[k, ] <- c(ii, j, r[ii, j, ])
      k <- k + 1
    }
  }
  r_2d <- r_2d %>%
    as.data.frame()
  colnames(r_2d)[1:2] <- c("iter", "chain")

  # R* calculation
  r1 <- r_2d %>% 
    arrange(chain, iter) %>% 
    select(-iter) %>% 
    mutate(chain=as.factor(chain))
  a_accuracy <- f_fit_gbm_predict(r1)

  # Split R* calculation
  half_rows <- iters / 2
  r_1h <- r_2d %>% 
    filter(iter<=half_rows)
  r_2h <- r_2d %>% 
    filter(iter>half_rows) %>% 
    mutate(chain=chain + chains)
  r2 <- bind_rows(r_1h, r_2h) %>% 
    select(-iter) %>% 
    mutate(chain=as.factor(chain))
  b_accuracy <- f_fit_gbm_predict(r2)

  res[[i]] <- cbind(iters, rep, rho, t(rs), "Rstar"=a_accuracy * n_distinct(r1$chain),
                    "Rstar_split"=b_accuracy * n_distinct(r2$chain))
}
res1 <- do.call(rbind, res)
saveRDS(res1, "../output/experiment_trends_joint_distribution.rds")
```

```{r}
g <- readRDS("../output/experiment_trends_joint_distribution.rds") %>%
  as.data.frame() %>% 
  select(iters, rho, rep, zsRhat, Rstar_split) %>% 
  unique() %>%
  select(-rep) %>% 
  melt(c("rho", "iters"))  %>% 
  mutate(variable=case_when(variable=="zsRhat"~"A. split-Rhat",
                            variable=="Rstar_split"~"B. split-R*")) %>% 
  ggplot(aes(y = value, x = rho)) + 
  geom_jitter() + 
  facet_grid(vars(iters), vars(variable)) + 
  geom_hline(yintercept = 1) +
  theme(strip.text.x = element_text(size=12, colour="black"),
        strip.text.y = element_text(size=12, colour="black"),
        axis.text.x = element_text(size=10, colour="black"),
        axis.text.y = element_text(size=10, colour="black")) +
  ylab("Value") +
  xlab(TeX("$\\rho$"))
g

ggsave("../output/trends_joint_distribution.pdf", g, width = 8, height = 6)
```

@@

ggplot(data = res, aes(x = rho, y = Rhat)) + 
  geom_jitter() + 
  facet_grid(vars(iters)) + 
  geom_hline(yintercept = 1) + 
  ggtitle('R* without splitting') +
  ylim(0, 6)

### 2020-03-27 write_up/ms.tex
added in trends examples

\subsection{Heterogeneity in chain variance: autoregressive example}\label{sec:heterogeneity}
We generate four Markov chains, where each samples from an autoregressive order 1 (AR1) process of the form,
%
\begin{equation}
X_t = \rho X_{t-1} + \epsilon_t
\end{equation}
%
where $\epsilon_t\stackrel{i.i.d.}{\sim}\mathcal{N}(0, \sigma)$ and $t=1,2,...,1000$. Three of the chains share the same $\sigma=1$, whereas the other chain has $\sigma=1/3$, so that it has $1/3$ of the (unconditional) variance of the others. In Fig. \ref{fig:ar1}A, we show how a GBM model fitted to these data classifies observations according to a sample's value. Unsurprisingly, since the fourth chain has a smaller variance, observations close to zero are likely to be classified as being generated by this chain.

@@

>>>>

\subsection{Heterogeneity in chain variance: autoregressive example}\label{sec:heterogeneity}
We generate four Markov chains, where each samples from an autoregressive order 1 (AR(1)) process of the form,
%
\begin{equation}
X_t = \rho X_{t-1} + \epsilon_t
\end{equation}
%
where $\epsilon_t\stackrel{i.i.d.}{\sim}\mathcal{N}(0, \sigma)$, $\rho=0.3$ and $t=1,2,...,1000$. Three of the chains share the same $\sigma=1$, whereas the other chain has $\sigma=1/3$, so that it has $1/3$ of the (unconditional) variance of the others. In Fig. \ref{fig:ar1}A, we show how a GBM model fitted to these data classifies observations according to a sample's value. Unsurprisingly, since the fourth chain has a smaller variance, observations close to zero are likely to be classified as being generated by this chain.

@@

\subsection{Non-stationary marginals}
To determine whether $R^*$ can detect non-stationary sampling distributions, we recapitulate an example from appendix A in \cite{vehtari2019rank}. This example showed that split-$\hat{R}$ could detect non-convergence caused by shifts in sampling distributions over time: in their case, they analysed chains with common linear trends in mean. Specifically, they first generated four chains by random sampling from a univariate normal distribution, then added a common time trend to each chain, resulting in a univariate distribution whose mean increased during sampling. We first repeat this analysis but using $R^*$ rather than $\hat{R}$: in Fig. \ref{fig:trends_all_dim}, we show these results. In column A, we show the results for $R^*$ calculated on the four chains that were run; column B shows the same calculation but after the chains are split into two equal halves. The different rows show the different sample sizes investigated: 250, 1000 and 4000; the horizontal axis shows the magnitude of time trend added to each sample; for all parameter sets, we run 10 replicates. This plot mirrors Fig. 18 in \cite{vehtari2019rank} and shows that, without splitting the chains, $R^*$ does not increase with trend whereas, after splitting, it does. As expected, split-$R^*$ is more reliably able to detect non-convergence as sample size increases.

These results are what would intuitively be expected: without systematic between-chain variation, it is not possible to reliably determine which of them caused a particular observation. In this case, because all chains exhibited the same secular trends over time, there would not be differences in their marginals. By splitting chains into two -- the first half being the early phase and, the second half being the later phase with higher mean -- this forces differences in the marginals. This meant it was possible to reliably pick whether an observation was caused by an early phase chain or a later one. As such, we recommend that $R^*$ always be calculated using split chains as is recommended for $\hat{R}$ \cite{carpenter2017stan,vehtari2019rank}.

\begin{figure}[!htb]
	\centerline{\includegraphics[width=1.0\textwidth]{../output/trends_all_dim.pdf}}
	\caption{\textbf{Univariate trends example.} Column A represents results for $R^*$ calculated using Algorithm \ref{alg:R_star} on the four chains; column B  shows the same calculation after each chain is split into two halves. The rows present the differing sample sizes. The horizontal axis measures (half) the change in mean across the whole sample: so a value ``1'' indicates the mean increases by 2 units from the start to end of sampling. At each parameter set, 10 replicates were run and jitter was added to the points. See the R markdown file on the Github repository for code to reproduce this example.}
	\label{fig:trends_all_dim}
\end{figure}

We next consider whether split-$R^*$ can detect non-convergence when only a single dimension trends. In Fig. \ref{fig:trends_one_dim}, we show the effect on $R^*$ of trends of different magnitudes across a range of target dimensions. In the simulations here, all dimensions bar one are stationary; the remaining dimension has a trend added to it. In all cases, split-$R^*$ increased with trend. Indeed, differences in the typical values of this metric were not apparent across the different target dimensions considered. This suggests split-$R^*$ can robustly determine chain identity if only a single dimension has a non-stationary sampling distribution.

\begin{figure}[!htb]
	\centerline{\includegraphics[width=1.0\textwidth]{../output/trends_one_dim.pdf}}
	\caption{\textbf{Multivariate trends example with split-$R^*$.} The columns represent different dimensionalities of the target distribution; the rows present different sample sizes. The horizontal axis measures (half) the change in mean across the single dimension that had a trend added to it: a value ``1'' indicates its mean increases by 2 units from the start to end of sampling; all other dimensions (if dimensions exceeded 1) had stationary distributions. At each parameter set, 10 replicates were run and jitter was added to the points. See the R markdown file on the Github repository for code to reproduce this example.}
	\label{fig:trends_one_dim}
\end{figure}

Means that trends over time is one form of non-stationarity; another is a time-varying covariance. Next, we consider a bivariate normal with (constant) standard normal marginals, but where the correlation between dimensions trends over time. Specifically, we allow the correlation to increase from $-\rho$ and $\rho$ throughout the course of simulations and use i.i.d. samples from the process across 4 ``chains''. Again, as before, $R^*$ calculated on unsplit chains is unable to detect this form of non-stationarity, since there are no inter-chain differences in the sampling distribution. Similarly, split-$\hat{R}$ does not detect this form of non-convergence since the marginal distribution across chains does not vary over time (Fig. \ref{fig:trends_joint_distribution}A). By contrast, split-$R^*$ can (Fig. \ref{fig:trends_joint_distribution}B), since it uses all information in the samples, including the covariance structure.

\begin{figure}[!htb]
	\centerline{\includegraphics[width=1.0\textwidth]{../output/trends_joint_distribution.pdf}}
	\caption{\textbf{Bivariate normal with trending correlation example.} Column A shows the results for split-$\hat{R}$; column B for $R^*$. The horizontal axis measures (half) the change in correlation across the whole sample: so a value ``0.5'' indicates the correlation increases by 1 unit (from -0.5 to 0.5) from the start to end of sampling. At each parameter set, 10 replicates were run and jitter was added to the points. See the R markdown file on the Github repository for code to reproduce this example.}
	\label{fig:trends_joint_distribution}
\end{figure}

A subtler form of non-stationarity can occur as the autocorrelation of a series increases. For example, if consider an AR(1) process, $X_t = \rho X_{t-1} + \epsilon_t$, where $\epsilon_t\stackrel{i.i.d.}{\sim}\mathcal{N}(0, \sigma)$. For $|\rho|<1$, the process has a stationary variance: $\sigma^2 / (1 - \rho^2)$. If $|\rho|=1$, however, the process becomes a random walk, with a variance that increases linearly with time. Any measure of convergence should, therefore, determine that the sampling distribution of an AR(1) process with $|\rho=1|$ is not converged since, by construction, such processes do not have a stationary target distribution. Any $\rho<1$ has a stationary distribution and, if sample sizes suffice,
